|

### 2023-01-09 paper_comparison/paper_comparison/paper_comparison.Rmd
major update

Diagnosing a model is the key to determining whether there is anything importantly wrong. For linear regression analysis, we typically interrogate the residuals for model diagnostics. Residuals summarise what is not captured by the model, and thus provide the capacity to identify what might be wrong.

We can assess residuals in multiple ways. Residuals might be plotted, as a histogram or quantile-quantile plot to examine the distribution. Using the classical normal linear regression model as an example, if the distribution is symmetric and unimodal, it is well-behaved. But if the distribution is skewed, bimodal, multimodal, or contains outliers, there is cause for concern. The distribution could also be inspected by conducting a goodness of fit test, such as the Shapiro-Wilk Normality test [@shapiro1965analysis].

More typically, residuals will be plotted, as a scatter plot against the predicted values and each of the explanatory variables to scrutinize their relationships. If there is any visually discoverable patterns, the model is potentially misspecified. However, correctly judging a residual plot where no pattern exists is a painstakingly difficult task for humans. It is especially common, particularly among new data analytics students to report patterns when an experienced data analyst might quickly conclude that there are none. Generally, one looks for noticeable departures from the model like non-linear dependency or heteroskedasticity. It is also possible to conduct hypothesis tests for non-linear dependence [@ramsey_tests_1969], and use a Breusch-Pagan test [@breusch_simple_1979] for heteroskedasticity.

Abundance of literature describe appropriate diagnostic methods for linear regression: @draper1998applied, @montgomery1982introduction, @belsley_regression_1980, @cook_applied_1999 and @cook1982residuals. The diligent reader of these sage writings will also notice sentences that express sentiments like *based on their experience, statistical tests are not widely used in regression diagnostics. The same or even larger amount of information can be provided by diagnostic plots than the corresponding tests in most empirical studies.* A common guidance by experts is that optimal method for diagnosing model fits is by plotting the data.

The persistence of this advice to check the plots is curious, and investigating why this might be common advice is the subject of this paper. The paper is structured as follows. The next background section describes the types of departures that one expects to detect, and describes a formal statistical process for reading residual plots, called visual inference. Section \ref{experiment-design} describes the experiment design to compare the decision made by formal hypothesis testing, and how humans would read diagnostic plots. The results are reported in Section \ref{results}. We finish with a discussion on future work, in particular how the responsibility for residual plot reading might be passed on to computer vision.

@@

```{r residual-plot-common-departures, fig.cap = "Example fitted vs residual plots: (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted plot."}

@@

  # Build a heter model with non-normal error
  # Reuse the x term
  mod$plot(heter_model(b = 0, e = rand_lognormal(sigma = 0.5))$
             gen(300, computed = select(dat, x)), 
    theme = theme_light()) + 
    ggtitle("D. Non-normality") +
    xlab("Fitted values") + 
    ylab("Residuals")
)
```

Graphical summaries in which residuals are plotted against fitted values or other functions of the regressors that are approximately orthogonal to residuals are referred to as standard residual plots in @cook1982residuals. As shown in Figure \ref{fig:residual-plot-common-departures}, the top-left panel is a good residual plot with residuals evenly distributed at both sides of the horizontal zero line showing no noticeable patterns.

There are various types of departures from a good residual plot. Non-linearity, heterskedasticity and non-normality are perhaps the three mostly checked departures.

@@

Compared to non-linearity and heteroskedasticity, non-normality is usually harder to detect from a residual plot since scatter plot do not readily reveal the marginal distribution. A favourable graphical summary for this task is the quantile-quantile plot. For a consistent comparison, the residual plot of this departure is still presented in plot D of Figure \ref{fig:residual-plot-common-departures}. However, since the scatter plot is not suitable for this task, non-normality will not be the focus of this paper. Besides, it is important to note that not all regression models assume normality for the error term, but a certain amount do including the classical normal linear regression model. In the case that the normality assumption is violated, it is expected to observe data points do not center around the horizontal axis and there is an uneven distribution of the number points at both below and above the horizontal axis. For example, given a skewed error distribution, fewer data points and more outliers are on one side of the horizontal axis as shown in plot D of Figure \ref{fig:residual-plot-common-departures}.

@@

Other than checking diagnostic plots, analysts may perform formal hypothesis testing for detecting model defects. Depending on the alternative hypothesis that is focused on, a variety of tests can be applied. For example, the presence of heteroskedasticity can usually be tested by applying the White test [@white_heteroskedasticity-consistent_1980] or the Breusch-Pagan test [@breusch_simple_1979], which are both derived from the Lagrange multiplier test [@silvey1959lagrangian] principle that relies on the asymptotic properties of the null distribution. For testing non-linearity, one may apply the F-test as a model structural test to examine the significance of specific polynomial and non-linear forms of the regressors, or the significance of proxy variables as in the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey_tests_1969]. And for testing normality, the Shapiro--Wilk test [@shapiro1965analysis] is perhaps the most widely used test included by many of the statistical softwares. Another choice will be the Jarque--Bera test [@jarque1980efficient] which directly checks if the sample skewness and kurtosis match a normal distribution.

Example residual plots given in Figure \ref{fig:residual-plot-common-departures} are examined by the corresponding RESET test, Breusch-Pagan test and Shapiro-Wilk test as shown in Table \ref{tab:example-residual-plot-table}. In the example, both the Breusch-Pagan test and the Shapiro--Wilk test rejects the null hypothesis $H_0$ for departures that they do not intend to examine. As discussed in @cook1982residuals, most residual-based tests for a particular type of departure from model assumptions are sensitive to other types of departures. It is likely $H_0$ is correctly rejected but for the wrong reason, which is known as the "Type III error". Additionally, outliers will often incorrectly trigger the rejection of $H_0$ despite when majority of the residuals are well-behaved [@cook_applied_1999]. Furthermore, with a sufficiently large sample size, residual-based tests may reject $H_0$ due to a slight departure that is of little practical significance. These can be largely avoided in diagnostic plots as experienced analysts can evaluate the acceptability of assumptions flexibly, even in the presence of outliers and slight departures.

@@

One may argue that reading diagnostic plots is to some extent subjective and indecisive compared to those rigorous statistical procedures as it relies on graphical perception - human's ability to interpret and decode the information embedded in the graph [@cleveland_graphical_1984]. It is particularly true that visual discovery suffers from its unsecured and unconfirmed nature where the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. One such example is finding an over-interpretation of the separation between gene groups in a two-dimensional projection from a linear discriminant analysis when in fact there are no differences in the expression levels between the gene groups and separation is not an uncommon occurrence [@roy_chowdhury_using_2015].

Visual inference was first introduced in a 1999 Joint Statistical Meetings (JSM) talk with the title "Inference for Data Visualization" by @buja_inference_1999 as an idea to address the issue of valid inference for visual discoveries of data plots [@gelman_exploratory_2004]. Later, @buja_statistical_2009 proposed the lineup protocol as a visual test inspired by the "police lineup" or "identity parade" which is the act of asking the eyewitness to identify criminal suspect from a group of irrelevant people. The protocol consists of $m$ randomly placed plots, where one plot is the actual data plot, and the remaining $m - 1$ plots have the identical graphical production as the data plot except the data has been replaced with data consistent with $H_0$. Then, an observer who have not seen the actual data plot will be asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the actual data plot would have no distinguishable difference with the null plots, and the probability of the observer correctly picks the actual data plot is $1/m$. If we reject $H_0$ as the observer correctly picks the actual data plot, then the Type I error of this test is $1/m$.

```{r first-example-lineup, fig.cap="Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data (plot $2^2 + 2$, exhibiting non-linearity) is embedded among 19 null plots, where the residuals were simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting plot $2^2 + 2$."}

@@

Figure \ref{fig:first-example-lineup} is an example of a lineup protocol. If the actual data plot at position $2^2 + 2$ is identifiable, then it is evidence for the rejection of $H_0$ that the regression model is correctly specified. In fact, the actual residual plot is obtained from a misspecified regression model with non-linearity defect.

The effectiveness of lineup protocol is validated by @majumder_validation_2013 under relatively simple classical normal linear regression model settings with one or two regressors. Their results suggest that visual test is capable of testing the significance of a single regressor with a similar power as a t-test, though they express that in general it is unnecessary to use visual inference if there exists a conventional test and they do not expect the visual test to perform equally well as the conventional test. In their third experiment, where there does not exist a proper conventional test, visual test outperforms the conventional test for a large margin. This is encouraging as it promotes the use of visual inference in border field of data science where there are no existing statistical testing procedures. Visual inference have also already been integrated into diagnostic of hierarchical linear models by @loy2013diagnostic, @loy2014hlmdiag and @loy2015you. They use lineup protocols to judge the assumption of linearity, normality and constant error variance for both the level-1 and level-2 residuals.

@@

Data used in the $m - 1$ null plots need to be simulated. In the context of regression diagnostics, sampling data from $H_0$ is equivalent to sampling data from the assumed model. As @buja_statistical_2009 suggested, $H_0$ is usually composited by a collection of distributions controlled by nuisance parameters. Since regression models can have various forms, there is no general solution to this problem, but it sometimes can be reduced to so called "reference distribution" by applying one of the three methods: (i) sampling from a conditional distribution given a minimal sufficient statistic under $H_0$, (ii) parametric bootstrap sampling with nuisance parameters estimated under $H_0$, and (iii) Bayesian posterior predictive sampling. The conditional distribution given a minimal sufficient statistic is the best justified reference distribution among the three [@buja_statistical_2009]. Essentially, null residuals can be simulated by regressing $N$ i.i.d standard normal random draws on the regressors, then rescaling it by the ratio of residual sum of square in two regressions.

@@

Further, a visual test can involve $K$ independent observers. Let $X_i = \{0,1\}$ be a binomial random variable denoting whether subject $i$ correctly detecting the actual data plot, and $X = \sum_{i=1}^{K}X_i$ be the number of observers correctly picking the actual data plot. Then, by imposing a relatively strong assumption on the visual test that all $K$ evaluations are fully independent, under $H_0$, $X \sim \mathrm{Binom}_{K,1/m}$. Therefore, the $p$-value of a lineup of size $m$ evaluated by $K$ observer is given as $P(X \geq x) = 1 - F(x) + f(x)$, where $F(.)$ is the cumulative distribution function, $f(.)$ is the probability mass function and $x$ is the realization of number of observers correctly picking the actual data plot [@majumder_validation_2013].

As pointed out by @vanderplas2021statistical, this basic binomial model doesn't take into account the possible dependencies in the visual test due to repeated evaluations of the same lineup. And it is inapplicable to visual test where subjects are asked to select one or more "most different" plots from the lineup. @vanderplas2021statistical summarises three common scenarios in visual inference: (1) $K$ different lineups are shown to $K$ subjects, (2) $K$ lineups with different null plots but the same actual data plot are shown to $K$ subjects, and (3) the same lineup is shown to $K$ subjects. Out of these three scenarios, Scenario 3 is the most common in previous studies as it puts the least constraints on the experiment design. For Scenario 3, @vanderplas2021statistical model the probability of a plot $i$ being selected from a lineup as $\theta_i$, where $\theta_i \sim Dirichlet(\alpha)$ for $i=1,...,m$ and $\alpha > 0$. And defined $c_i$ to be the number of times plot $i$ being selected in $K$ evaluations. In case subject $j$ makes multiple selections, $1/s_j$ will be added to $c_i$ instead of one, where $s_j$ is the number of plots subject $j$ selected for $j=1,...K$. This ensured $\sum_{i}c_i=K$. Since we are only interested in the selections of the actual data plot $i$, the marginal model can be simplified to a beta-binomial model and thus the visual p-value is given as

@@

Note that Equation \ref{eq:pvalue-beta-binomial} only works with non-negative integer $c_i$. For non-negative real number $c_i$, linear approximation will be applied to calculate the p-value

@@

\begin{equation}
P(C = c_i) = {K \choose c_i}\frac{B(c_i + \alpha, K - c_i + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+.
\end{equation}

Besides, the parameter $\alpha$ used in Equation \ref{eq:pvalue-beta-binomial} is usually unknown and hence needs to be estimated from the survey data. For low values of $\alpha$, only a few plots are attractive to the observers and tend to be selected. For higher values of $\alpha$, the distribution of the probability of each plot being selected is more even. @vanderplas2021statistical define that a plot is $c$-interesting if $c$ or more participants select the plot as the most different. Given the definition, The expected number of plots selected at least $c$ times, $E[Z_c]$, is calculated as

@@

@vanderplas2021statistical suggest that $\alpha$ can be estimated using MLE with Equation \ref{eq:c-interesting-expectation}. But for precise estimate of $\alpha$, additional responses to Rorschach lineups, which is a type of lineup that consists of plots constructed from the same null data generating mechanism, are required.

@@

As discussed in @majumder_validation_2013, individual's skill will affect the number of observers who identify the actual data plot from the lineup. Thus, the power of a visual test depends on the subject-specific abilities. Previously, it is addressed by modelling the probability of a subject $i$ correctly picking the actual data plot from a lineup $l$ using a mixed-effect logistic regression with the subject being treated as a random effect [@majumder_validation_2013]. However, having this probability is insufficient to determine the power of a visual test allowed for multiple selections as it doesn't provide information about the number of selections made by the subject for p-value calculation.

Instead, we omitted the individual effect and directly estimated the probability of a lineup being rejected using a logistic regression with the natural logarithm of the effect size as the only regressor formulated as

```{=tex}
\begin{equation} \label{eq:logistic-regression-1-1}
Pr(\text{reject}~H_0|H_1,E) = \Lambda(\beta_0 + \beta_1 log_e(\boldsymbol{E})),
\end{equation}
```
\noindent where $\Lambda(.)$ is the standard logistic function given as $\Lambda(z) = exp(z)/(1+exp(z))$.

@@

Three experiments were conducted to investigate the difference between conventional hypothesis testing and visual inference in the application of linear regression diagnostics. The experiment I has ideal scenario for conventional testing, where the visual test is not expected to outperform the conventional test. Meanwhile, the experiment II is a scenario where the conventional test is an approximate test, in which the visual test may have a chance to match the performance of the conventional test. The experiment III is designed for collecting human responses to lineup with only good residual plots such that the parameter $\alpha$ in Equation \ref{eq:pvalue-beta-binomial} can be estimated. Overall, we planned to collect 7974 evaluations on 1152 uniqued lineups performed by 443 subjects throughout three experiment.

@@

Two types of departures, namely non-linearity and heteroskedasticity, were considered with the corresponding data generating process being designed for experiment I and II.

@@

(include English reference)

@@

\begin{align} \label{eq:nonlinearity-model}
\boldsymbol{y} = 1 + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} = g(\boldsymbol{x}_{raw}, 1), \\
\boldsymbol{z} = g(\boldsymbol{z}_{raw}, 1), \\
\boldsymbol{z}_{raw} = He_j(g(\boldsymbol{z}, 2)),
\end{align}

\noindent where $\boldsymbol{y}$, $\boldsymbol{x}$, $\boldsymbol{\varepsilon}$, $\boldsymbol{x}_{raw}$, $\boldsymbol{z}_{raw}$ are vectors of size $n$, $He_{j}(.)$ is the $j$th-order probabilist's Hermite polynomials, $\varepsilon \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)$, and $g(\boldsymbol{x}, k)$ is a scaling function to enforce the support of the random vector to be $\{-k, k\}$ defined as

@@

Hermite polynomials were initially defined by @de1820theorie, but named after Hermite [@hermite1864nouveau] because of the unrecognisable form of Laplace's work. When simulating $Z_{raw}$, the function `hermite` from the R package `mpoly` [@mpoly] was used to generate Hermite polynomials.

@@

Since $z = O(x^j)$, for $j > 1$, $z$ is a higher order term leaves out by the null regression, which will result in model misspecification. Visual patterns of non-linearity were simulated using four different order of probabilist's Hermite polynomials ($j = 2, 3, 6, 18$) and four different distribution of $X_{raw}$: (1) $U(-1, 1)$, (2) $N(0, 0.3^2)$, (3) $lognormal(0, 0.6^2)/3$ and (4) $u\{1, 5\}$. A summary of the factors is given in Table \ref{tab:model-factor-table}.

@@

Figure \ref{fig:example-poly-lineup} shows one of the lineups used in experiment I. This lineup is produced under $j = 6$ and $X_{raw} \sim N(0.0.3^2)$. The actual data plot location is $2^3 - 4$. All five subjects correctly identified the actual data plot for this lineup.

>>>>

Diagnosing a model is the key to determining whether there is anything importantly wrong. In linear regression analysis, residuals are typically examined for model diagnostics. Residuals summarise what is not captured by the model, and thus provide the capacity to identify what might be wrong.

One can access residuals in multiple ways. Residuals might be plotted, as a histogram or quantile-quantile plot to examine the distribution. Using the classical normal linear regression model as an example, if the distribution is symmetric and unimodal, it is well-behaved. But if the distribution is skewed, bimodal, multimodal, or contains outliers, there is cause for concern. One could also inspect the distribution by conducting a goodness of fit test, such as the Shapiro-Wilk Normality test [@shapiro1965analysis].

More typically, residuals will be plotted, as a scatter plot against the predicted values and each of the explanatory variables to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially misspecified. In general, one looks for noticeable departures from the model like non-linear dependency or heteroskedasticity. However, correctly judging a residual plot where no pattern exists is a painstakingly difficult task for humans (?citation). It is especially common, particularly among new data analysts, to report patterns when an experienced one might quickly conclude that there are none. It is also possible to conduct hypothesis tests for non-linear dependence [@ramsey_tests_1969], and use a Breusch-Pagan test [@breusch_simple_1979] for heteroskedasticity.

Abundance of literature describe appropriate diagnostic methods for linear regression: @draper1998applied, @montgomery1982introduction, @belsley_regression_1980, @cook_applied_1999 and @cook1982residuals. The diligent reader of these sage writings will also notice sentences that express sentiments like *based on their experience, statistical tests are not widely used in regression diagnostics since the same or even larger amount of information can be provided by diagnostic plots than the corresponding tests in most empirical studies.* A common guidance by experts is that optimal method for diagnosing model fits is by plotting the data.

The persistence of this advice to check the plots is curious, and investigating why this might be common advice is the subject of this paper. The paper is structured as follows. The next background section describes the types of departures that one expects to detect, and outlines a formal statistical process for reading residual plots, called visual inference. Section \ref{experimental-design} details the experimental design to compare the decision made by formal hypothesis testing, and how humans would read diagnostic plots. The results are reported in Section \ref{results}. We finish with a discussion on future work, in particular how the responsibility for residual plot reading might be relieved.

@@

```{r residual-plot-common-departures, fig.width = 10, fig.height = 2.5, fig.cap = "Example fitted vs residual plots: (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted plot."}

@@

  # Build a heter model with non-normal error
  # Reuse the x term
  mod$plot(heter_model(b = 0, e = rand_lognormal(sigma = 0.5))$
             gen(300, computed = select(dat, x)), 
    theme = theme_light()) + 
    ggtitle("D. Non-normality") +
    xlab("Fitted values") + 
    ylab("Residuals"),
  ncol = 4
)
```

Graphical summaries in which residuals are plotted against fitted values or other functions of the regressors that are approximately orthogonal to residuals are referred to as standard residual plots in @cook1982residuals. As shown in Figure \ref{fig:residual-plot-common-departures}, the plot A is an ideal residual plot with residuals evenly distributed at both sides of the horizontal zero line, with no noticeable patterns.

There are various types of departures from an ideal residual plot. Non-linearity, heterskedasticity and non-normality are perhaps the three mostly checked departures.

@@

Compared to non-linearity and heteroskedasticity, non-normality is usually harder to detect from a residual plot since a scatter plot do not readily reveal the marginal distribution. A favourable graphical summary for this task is the quantile-quantile plot. Hence, non-normality will not be the focus of this paper as we mainly discuss residual plots. For a consistent comparison, the residual plot of this departure is still presented in plot D of Figure \ref{fig:residual-plot-common-departures}. Besides, it is important to note that not all regression models assume normality for the error term, but a certain amount do including the classical normal linear regression model. In the case that the normality assumption is violated, it is expected to observe data points do not center around the horizontal axis and there is an uneven distribution of the number points at both below and above the horizontal axis. For example, given a skewed error distribution, there will be fewer data points and more outliers on one side of the horizontal axis as shown in plot D of Figure \ref{fig:residual-plot-common-departures}.

@@

Other than checking diagnostic plots, analysts may perform formal hypothesis testing for detecting model defects. Depending on the alternative hypothesis that is focused on, a variety of tests can be applied. For example, the presence of heteroskedasticity can usually be tested by applying the White test [@white_heteroskedasticity-consistent_1980] or the Breusch-Pagan test [@breusch_simple_1979], which are both derived from the Lagrange multiplier test [@silvey1959lagrangian] principle that relies on the asymptotic properties of the null distribution. For testing non-linearity, one may apply the F-test as a model structural test to examine the significance of specific polynomial and non-linear forms of the regressors, or the significance of proxy variables as in the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey_tests_1969]. The Shapiro-Wilk test [@shapiro1965analysis] is the most widely used test of non-normality included by many of the statistical softwares. The Jarque--Bera test [@jarque1980efficient] is also used to directly checks if the sample skewness and kurtosis match a normal distribution.

Example residual plots given in Figure \ref{fig:residual-plot-common-departures} are examined by the corresponding RESET test, Breusch-Pagan test and Shapiro-Wilk test as shown in Table \ref{tab:example-residual-plot-table}. In the example, the Breusch-Pagan test and the Shapiro--Wilk test both reject the null hypothesis $H_0$ for departures that they do not intend to examine. As discussed in @cook1982residuals, most residual-based tests for a particular type of departure from model assumptions are also sensitive to other types of departures. It is likely $H_0$ is correctly rejected but for the wrong reason, a phenomenon known as the "Type III error". Additionally, outliers will often incorrectly trigger the rejection of $H_0$ despite when majority of the residuals are well-behaved [@cook_applied_1999]. Furthermore, with a sufficiently large sample size, residual-based tests may reject $H_0$ due to a slight departure that is of little practical significance. These can be largely avoided in diagnostic plots as experienced analysts can evaluate the acceptability of assumptions flexibly, even in the presence of outliers and slight departures.

@@

One may argue that reading diagnostic plots is to some extent subjective and indecisive compared to those rigorous statistical procedures as it relies on graphical perception - human ability to interpret and decode the information embedded in graph [@cleveland_graphical_1984]. Further, the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. For instance, people 
over-interpret the separation between gene groups in a two-dimensional projection from a linear discriminant analysis when in fact there are no differences in the expression levels between the gene groups and separation is not an uncommon occurrence [@roy_chowdhury_using_2015].

Visual inference was first introduced in a 1999 Joint Statistical Meetings (JSM) talk with the title "Inference for Data Visualization" by @buja_inference_1999 as an idea to address the issue of valid inference for visual discoveries of data plots. Later, @buja_statistical_2009 proposed the lineup protocol as a visual test inspired by the "police lineup" or "identity parade" which is the act of asking the eyewitness to identify criminal suspect from a group of irrelevant people. The protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ plots have the identical graphical procedure except the data has been replaced with data consistent with $H_0$. Then, an observer who have not seen the data plot will be asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$.

```{r first-example-lineup, fig.height = 5, fig.cap="Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data (plot $2^2 + 2$, exhibiting non-linearity) is embedded among 19 null plots, where the residuals are simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting plot $2^2 + 2$."}

@@

Figure \ref{fig:first-example-lineup} is an example of a lineup protocol. If the data plot at position $2^2 + 2$ is identifiable, then it is evidence for the rejection of $H_0$ that the regression model is correctly specified. In fact, the actual residual plot is obtained from a misspecified regression model with non-linearity defect.

The effectiveness of lineup protocol for regression analysis is validated by @majumder_validation_2013 under relatively simple settings with up to two regressors. Their results suggest that visual tests are capable of testing the significance of a single regressor with a similar power as a t-test, though they express that in general it is unnecessary to use visual inference if there exists a conventional test, and they do not expect the visual test to perform equally well as the conventional test. In their third experiment, where there is not a conventional test, visual test outperforms the conventional test for a large margin. This is encouraging, as it promotes the use of visual inference in situations where there are no existing statistical testing procedures. Visual inference have also been integrated into diagnostic of hierarchical linear models by @loy2013diagnostic, @loy2014hlmdiag and @loy2015you. They use lineup protocols to judge the assumption of linearity, normality and constant error variance for both the level-1 and level-2 residuals. (expand?)

@@

Data used in the $m - 1$ null plots needs to be simulated. In the context of regression diagnostics, sampling data from $H_0$ is equivalent to sampling data from the assumed model. As @buja_statistical_2009 suggested, $H_0$ is usually a composite hypothesis controlled by nuisance parameters. Since regression models can have various forms, there is no general solution to this problem, but it sometimes can be reduced to a so called "reference distribution" by applying one of the three methods: (i) sampling from a conditional distribution given a minimal sufficient statistic under $H_0$, (ii) parametric bootstrap sampling with nuisance parameters estimated under $H_0$, and (iii) Bayesian posterior predictive sampling. The conditional distribution given a minimal sufficient statistic is the best justified reference distribution among the three [@buja_statistical_2009]. Essentially, null residuals can be simulated by regressing $N$ i.i.d standard normal random draws on the regressors, then rescaling it by the ratio of residual sum of square in two regressions.

@@

In hypothesis testing, $p$-value is defined as the probability of observing test results as least as extreme as the observed result given $H_0$ is true. Within the context of visual inference, by involving $k$ independent observers, $p$-value can be interpreted as the probability of having as many or more subjects detect the data plot than the observed result.

Let $X_j = \{0,1\}$ be a binomial random variable denoting whether subject $j$ correctly detecting the data plot, and $X = \sum_{j=1}^{K}X_j$ be the number of observers correctly picking the data plot. Then, by imposing a relatively strong assumption on the visual test that all $K$ evaluations are fully independent, under $H_0$, $X \sim \mathrm{Binom}_{K,1/m}$. Therefore, the $p$-value of a lineup of size $m$ evaluated by $K$ observer is given as $P(X \geq x) = 1 - F(x) + f(x)$, where $F(.)$ is the cumulative distribution function, $f(.)$ is the probability mass function and $x$ is the realization of number of observers correctly picking the data plot [@majumder_validation_2013].

As pointed out by @vanderplas2021statistical, this basic binomial model doesn't take into account the possible dependencies in the visual test due to repeated evaluations of the same lineup. And it is inapplicable to visual test where subjects are asked to select one or more "most different" plots from the lineup. @vanderplas2021statistical summarises three common scenarios in visual inference: (1) $K$ different lineups are shown to $K$ subjects, (2) $K$ lineups with different null plots but the same data plot are shown to $K$ subjects, and (3) the same lineup is shown to $K$ subjects. Out of these three scenarios, Scenario 3 is the most common in previous studies as it puts the least constraints on the experiment design. For Scenario 3, @vanderplas2021statistical models the probability of a plot $i$ being selected from a lineup as $\theta_i$, where $\theta_i \sim Dirichlet(\alpha)$ for $i=1,...,m$ and $\alpha > 0$. The number of times plot $i$ being selected in $K$ evaluations is denoted as $c_i$. In case subject $j$ makes multiple selections, $1/s_j$ will be added to $c_i$ instead of one, where $s_j$ is the number of plots subject $j$ selected for $j=1,...K$. This ensures $\sum_{i}c_i=K$. Since we are only interested in the selections of the data plot $i$, the marginal model can be simplified to a beta-binomial model and thus the visual p-value is given as

@@

Note that Equation \ref{eq:pvalue-beta-binomial} given in @vanderplas2021statistical only works with non-negative integer $c_i$. We extend the equation to non-negative real number $c_i$ by applying a linear approximation

@@

\begin{equation} \label{eq:pmf-beta-binomial}
P(C = c_i) = {K \choose c_i}\frac{B(c_i + \alpha, K - c_i + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+.
\end{equation}

Besides, the parameter $\alpha$ used in Equation \ref{eq:pvalue-beta-binomial} and \ref{eq:pmf-beta-binomial} is usually unknown and hence needs to be estimated from the survey data. For low values of $\alpha$, only a few plots are attractive to the observers and tend to be selected. For higher values of $\alpha$, the distribution of the probability of each plot being selected is more even. @vanderplas2021statistical defines that a plot is $c$-interesting if $c$ or more participants select the plot as the most different. Given the definition, The expected number of plots selected at least $c$ times, $E[Z_c]$, is calculated as

@@

With Equation \ref{eq:c-interesting-expectation}, $\alpha$ can be estimated using maximum likelihood estimation. But for precise estimate of $\alpha$, additional responses to Rorschach lineups, which is a type of lineup that consists of plots constructed from the same null data generating mechanism, are required.

@@

The power of a model misspecification test specifies the chance of a misspecified regression model being detected given a certain alternative hypothesis. It is an important indicator when one concerns if the regression model is misspecified. Though, people might usually be more interested in knowing how much the residuals depart from the model assumptions and if the departure is of practical significance.

As discussed in @majumder_validation_2013, individual's skill may affect the number of observers who identify the data plot from the lineup. Thus, the power of a visual test could depend on the subject-specific abilities. Previously, it is addressed by modelling the probability of a subject $j$ correctly picking the data plot from a lineup $l$ using a mixed-effect logistic regression with the subject being treated as a random effect [@majumder_validation_2013]. However, in the multiple selections scenario, having this probability is insufficient to determine the power of a visual test as it doesn't provide information about the number of selections made by the subject for p-value calculation.

Instead, we estimate the probability of a lineup being rejected directly by assuming the individual skill has negligible impact on the variation of the power of a lineup. The assumption is made to simplify the model structure, thus mitigating the need for expensive large-scale experiments to estimate the complex covariance matrix. The model is a logistic regression with the natural logarithm of the effect as the only regressor formulated as

```{=tex}
\begin{equation} \label{eq:logistic-regression-1-1}
Pr(\text{reject}~H_0|H_1,E) = \Lambda(\beta_0 + \beta_1 log_e(E)),
\end{equation}
```
\noindent where $\Lambda(.)$ is the standard logistic function given as $\Lambda(z) = exp(z)/(1+exp(z))$.

```{r eval = FALSE}
mix_effect_dat <- vi_survey %>%
  filter(!attention_check) %>%
  filter(!null_lineup) %>%
  filter(x_dist == "uniform") %>%
  filter(type == "polynomial") %>%
  mutate(log_effect_size = log(effect_size)) %>%
  mutate(subject = factor(as.integer(factor(paste(exp, set))))) %>%
  mutate(reject = p_value < 0.05) %>%
  select(log_effect_size, reject, subject)

# Singularity issue (random effect too small? not worth to fit a mix-effect model?)
mix_mod <- lme4::glmer(reject ~ scale(log_effect_size) + (1 | subject), data = mix_effect_dat, family = binomial())

pred_dat <- map_df(seq(-1, 6, 0.1),
    function(x) {
      result <- data.frame(log_effect_size = rep(x, length(unique(mix_effect_dat$subject))))
      pred <- predict(mix_mod, 
                      type = "response", 
                      newdata = data.frame(log_effect_size = x, 
                                           subject = unique(mix_effect_dat$subject)))
      result$subject <- names(pred)
      result$pred <- pred
      result
    })

pred_dat %>%
  ggplot() +
  geom_line(aes(log_effect_size, pred, group = subject))
```

@@

Three experiments are conducted to investigate the difference between conventional hypothesis testing and visual inference in the application of linear regression diagnostics. The experiment I has ideal scenario for conventional testing, where the visual test is not expected to outperform the conventional test. Meanwhile, the experiment II is a scenario where the conventional test is an approximate test, in which the visual test may have a chance to match the performance of the conventional test. The experiment III is designed for collecting human responses to null lineups such that the parameter $\alpha$ in Equation \ref{eq:pvalue-beta-binomial} can be estimated. Overall, we plan to collect 7974 evaluations on 1152 uniqued lineups performed by 443 subjects throughout three experiments.

@@

Two types of departures, namely non-linearity and heteroskedasticity, are considered with the corresponding data generating process being designed for experiment I and II.

@@

@@

\begin{align} \label{eq:nonlinearity-model}
\boldsymbol{y} = 1 + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} = g(\boldsymbol{x}_{raw}, 1), \\
\boldsymbol{z} = g(\boldsymbol{z}_{raw}, 1), \\
\boldsymbol{z}_{raw} = He_j(g(\boldsymbol{x}, 2)),
\end{align}

\noindent where $\boldsymbol{y}$, $\boldsymbol{x}$, $\boldsymbol{\varepsilon}$, $\boldsymbol{x}_{raw}$, $\boldsymbol{z}_{raw}$ are vectors of size $n$, $He_{j}(.)$ is the $j$th-order probabilist's Hermite polynomials, $\varepsilon \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)$, and $g(\boldsymbol{x}, k)$ is a scaling function to enforce the support of the random vector to be $[-k, k]^n$ defined as

@@

According to @abramowitz1964handbook, Hermite polynomials were initially defined by @de1820theorie, but named after Hermite [@hermite1864nouveau] because of the unrecognisable form of Laplace's work. When simulating $\boldsymbol{z}_{raw}$, function `hermite` from the R package `mpoly` [@mpoly] is used to generate Hermite polynomials.

@@

Since $z = O(x^j)$, for $j > 1$, $z$ is a higher order term leaves out by the null regression, which will lead to model misspecification.

Visual patterns of non-linearity are simulated using four different orders of probabilist's Hermite polynomials ($j = 2, 3, 6, 18$) and four different distributions of $X_{raw}$: (1) $U(-1, 1)$, (2) $N(0, 0.3^2)$, (3) $lognormal(0, 0.6^2)/3$ and (4) $u\{1, 5\}$. A summary of the factors is given in Table \ref{tab:model-factor-table}.

@@

Figure \ref{fig:example-poly-lineup} demonstrates one of the lineups used in experiment I. This lineup is produced by the non-linearity model under $j = 6$ and $X_{raw} \sim N(0.0.3^2)$. The data plot location is $2^3 - 4$. All five subjects correctly identify the data plot from this lineup.
